---
title: "QC validation & annotation panels (reviewer response)"
author: "CPS"
date: "2026-06-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.path = "../../results/figures/", dev = "png", dpi = 300)
options(future.globals.maxSize = 1e11)
Sys.setenv(R_MAX_VSIZE = "100Gb")
```

Panels for reviewer concerns:
- **#1 / #2** QC metrics across axes of variability (Reviewer #3.1)
- **#5** Annotation validation — dot plot / stacked violin, incl. myeloid subsets (Reviewer #1.2, #1-minor-2, #3.2)
- **#6** Cluster/condition mixing — composition + **iLISI/cLISI** mixing scores (Reviewer #2)

All figures are written to `manuscript/figures/` by the `fig.path` setting above.

```{r libs}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Load data
```{r load}
obj <- readRDS("../../data/processed/Emouse.rds")
DefaultAssay(obj) <- "RNA"

# condition label (genotype x sex x age = 12 groups) and ordered factors
obj$group <- paste(obj$genotype, obj$sex, obj$age, sep = "_")
obj$age   <- factor(obj$age, levels = c("06Mo", "12Mo", "18Mo"))

# use the curated cell-type label; drop the technical "negative"/"hybrid" tags for marker panels
ct_col <- "celltype"
Idents(obj) <- ct_col

dir.create("../../results/figures", showWarnings = FALSE, recursive = TRUE)
qc <- c("nCount_RNA", "nFeature_RNA", "percent.mt", "log10GenesPerUMI")
```

---

## 1. QC metrics across conditions (Reviewer #3.1)

### 1a. Per-sample violins (47 samples) — sample-to-sample consistency
```{r QC_per_sample_violin, fig.width = 14, fig.height = 11}
p_ps <- VlnPlot(obj, features = qc, group.by = "orig.ident", pt.size = 0, ncol = 1) &
  theme(axis.text.x = element_text(size = 5, angle = 90, hjust = 1),
        axis.title.x = element_blank())
p_ps
```

### 1b. QC metrics split by each axis of variability
```{r QC_by_axis_violin, fig.width = 14, fig.height = 10}
mk_axis <- function(grp) {
  VlnPlot(obj, features = qc, group.by = grp, pt.size = 0, ncol = 4) &
    theme(plot.title = element_text(size = 9))
}
(mk_axis("age") / mk_axis("sex") / mk_axis("genotype"))
```

### 1c. Joint QC scatter (covariates considered jointly)
```{r QC_joint_scatter, fig.width = 12, fig.height = 5}
s1 <- FeatureScatter(obj, "nCount_RNA", "nFeature_RNA", group.by = "orig.ident") +
  NoLegend() + scale_x_log10() + scale_y_log10() + ggtitle("Depth vs complexity")
s2 <- FeatureScatter(obj, "nCount_RNA", "percent.mt", group.by = "age") +
  scale_x_log10() + ggtitle("Depth vs % mito")
s1 | s2
```

---

## 2. Cells recovered & QC distribution per sample (Reviewer #3.1)

### 2a. Nuclei recovered per sample, faceted by age, colored by genotype x sex
```{r cells_per_sample, fig.width = 13, fig.height = 6}
cell_counts <- obj@meta.data %>%
  count(orig.ident, genotype, sex, age, name = "nuclei")

ggplot(cell_counts, aes(orig.ident, nuclei, fill = interaction(genotype, sex))) +
  geom_col() +
  facet_grid(~ age, scales = "free_x", space = "free") +
  scale_fill_brewer(palette = "Set2", name = "genotype.sex") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 6, hjust = 1, vjust = 0.5)) +
  labs(x = NULL, y = "nuclei recovered")
```

### 2b. Median QC per sample (table-as-figure) to show no condition-driven drift
```{r QC_summary_heatmap, fig.width = 11, fig.height = 7}
qc_summary <- obj@meta.data %>%
  group_by(orig.ident, genotype, sex, age) %>%
  summarise(across(all_of(qc), median), nuclei = n(), .groups = "drop")

qc_long <- tidyr::pivot_longer(qc_summary, all_of(qc),
                               names_to = "metric", values_to = "median_value")

ggplot(qc_long, aes(orig.ident, metric, fill = median_value)) +
  geom_tile() +
  geom_text(aes(label = round(median_value, 2)), size = 2) +
  facet_grid(metric ~ age, scales = "free", space = "free_x") +
  scale_fill_viridis_c() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 6, hjust = 1, vjust = 0.5),
        axis.text.y = element_blank(), strip.text.y = element_text(size = 7)) +
  labs(x = NULL, y = NULL, fill = "per-sample\nmedian")
```

---

## 5. Annotation validation — canonical markers (Reviewer #1.2, #1-minor-2, #3.2)

Microglia are resolved beyond `Csf1r` alone, and macrophage/monocyte markers are
included so the myeloid compartment can be evaluated (Reviewer #1.2).

```{r marker_panels}
markers <- list(
  Excitatory  = c("Snap25", "Slc17a7", "Rbfox3"),
  Inhibitory  = c("Gad1", "Gad2"),
  Astrocyte   = c("Aqp4", "Gfap", "Slc1a3"),
  Oligo       = c("Plp1", "Mbp", "Mog"),
  OPC         = c("Pdgfra", "Olig1", "Olig2"),
  Microglia   = c("Csf1r", "Ctss", "Tmem119", "P2ry12", "Cx3cr1"),
  Macrophage  = c("Mrc1", "Lyve1", "Cd163", "Ms4a7"),
  Monocyte    = c("Ly6c2", "Ccr2"),
  Endothelial = c("Cldn5", "Flt1", "Pecam1"),
  Mural       = c("Pdgfrb", "Rgs5", "Kcnj8", "Acta2"))

# restrict to biologically annotated clusters via idents (no object copy -> memory safe)
Idents(obj) <- "celltype"
keep_ct <- setdiff(levels(factor(obj$celltype)), c("negative", "hybrid"))
```

### 5a. Dot plot
```{r annot_dotplot, fig.width = 15, fig.height = 6}
DotPlot(obj, features = markers, idents = keep_ct, cluster.idents = TRUE) +
  RotatedAxis() +
  theme(axis.text.x = element_text(size = 7), strip.text = element_text(size = 8)) +
  labs(x = NULL, y = NULL)
gc()
```

### 5b. Stacked violin
```{r annot_stacked_violin, fig.width = 11, fig.height = 13}
VlnPlot(obj, features = unlist(markers, use.names = FALSE),
        idents = keep_ct, stack = TRUE, flip = TRUE, pt.size = 0) +
  NoLegend() +
  theme(axis.text.x = element_text(size = 8), axis.text.y = element_text(size = 7))
gc()
```

---

## 6. Cluster / condition mixing (Reviewer #2)

### 6a. UMAP by cell type, split by genotype
```{r umap_split_genotype, fig.width = 12, fig.height = 6}
DimPlot(obj, group.by = "celltype", split.by = "genotype",
        raster = TRUE, raster.dpi = c(600, 600)) +
  ggtitle("Cell type by genotype")
```

### 6b. UMAP colored by sample (batch overlay)
```{r umap_by_sample, fig.width = 8, fig.height = 6}
DimPlot(obj, group.by = "orig.ident", raster = TRUE, raster.dpi = c(600, 600)) +
  NoLegend() + ggtitle("Sample overlay (no condition-specific island)")
```

### 6c. Per-sample composition (quantitative, not eyeballed)
```{r composition_per_sample, fig.width = 13, fig.height = 6}
comp <- obj@meta.data %>% count(orig.ident, age, celltype)
ggplot(comp, aes(orig.ident, n, fill = celltype)) +
  geom_col(position = "fill") +
  facet_grid(~ age, scales = "free_x", space = "free") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 6, hjust = 1, vjust = 0.5)) +
  labs(x = NULL, y = "proportion of nuclei")
```

### 6d. LISI mixing scores (Reviewer #2 — quantitative batch metric)

We use the Local Inverse Simpson's Index (Korsunsky et al. 2019). For each nucleus,
neighbours in PCA space are Gaussian-weighted to a fixed perplexity and the inverse
Simpson index of a label is taken — the **effective number of categories** in the
local neighbourhood.

- **iLISI** on `orig.ident` (sample/batch): ranges 1 (no mixing) → 47 (all samples
  equally represented locally). **Higher = better mixed.**
- **cLISI** on `celltype`: ideal ≈ 1 (neighbourhoods are cell-type-pure).

A self-contained R implementation is used so the result does not depend on the
`lisi` package compiling (its pinned C++11 flag conflicts with current toolchains).

```{r lisi_fun}
# Pure-R LISI (Korsunsky et al. 2019). Returns effective #categories per cell.
compute_lisi <- function(emb, labels, perplexity = 30) {
  labels <- as.factor(labels)
  k   <- perplexity * 3
  knn <- FNN::get.knn(emb, k = k)
  D   <- knn$nn.dist^2
  idx <- knn$nn.index
  n   <- nrow(emb); logU <- log(perplexity); out <- numeric(n)
  for (i in seq_len(n)) {
    Di <- D[i, ]; beta <- 1; betamin <- -Inf; betamax <- Inf
    for (t in 1:50) {                              # Hbeta binary search (as in t-SNE)
      P <- exp(-Di * beta); sumP <- sum(P)
      if (sumP == 0) { P <- rep(0, k); H <- 0 } else {
        H <- log(sumP) + beta * sum(Di * P) / sumP; P <- P / sumP }
      if (abs(H - logU) < 1e-5) break
      if (H > logU) { betamin <- beta; beta <- if (is.infinite(betamax)) beta*2 else (beta+betamax)/2 }
      else          { betamax <- beta; beta <- if (is.infinite(betamin)) beta/2 else (beta+betamin)/2 }
    }
    sp <- tapply(P, labels[idx[i, ]], sum); sp[is.na(sp)] <- 0
    out[i] <- 1 / sum(sp^2)                         # inverse Simpson
  }
  out
}
```

The naive ceiling for iLISI is the number of samples (47), but that is unreachable:
with a perplexity-30 neighbourhood (~30–90 cells) the inverse-Simpson index cannot
represent 47 samples even if labels were random. The correct reference is therefore
the **label-shuffled null** — iLISI recomputed on the *same* embedding after
permuting `orig.ident` — which is the maximum mixing achievable on this graph. The
floor is ~1 (total segregation).

LISI on all 555k nuclei takes ~20 min per pass, so results are cached to
`results/figures/`. Delete those `.rds` files to force a recompute.

```{r lisi_compute}
library(FNN)
emb_s     <- Embeddings(obj, "pca")[, 1:15]
val_cache  <- "../../results/figures/lisi_values_all_nuclei.rds"  # observed iLISI/cLISI/celltype
null_cache <- "../../results/figures/ilisi_null_all_nuclei.rds"   # shuffled-null iLISI

# --- observed (load cache or compute on ALL nuclei) ---
if (file.exists(val_cache)) {
  ldat0 <- readRDS(val_cache)
  iLISI <- ldat0$iLISI; cLISI <- ldat0$cLISI
} else {
  iLISI <- compute_lisi(emb_s, obj$orig.ident)
  cLISI <- compute_lisi(emb_s, obj$celltype)
  saveRDS(data.frame(iLISI = iLISI, cLISI = cLISI, celltype = obj$celltype), val_cache)
}

# --- perfect-mixing null = achievable ceiling (load cache or compute) ---
if (file.exists(null_cache)) {
  iLISI_null <- readRDS(null_cache)
} else {
  set.seed(1)
  iLISI_null <- compute_lisi(emb_s, sample(obj$orig.ident))
  saveRDS(iLISI_null, null_cache)
}

n_samp    <- length(unique(obj$orig.ident))
null_ceil <- median(iLISI_null)                # reference line for the boxplot
ldat <- data.frame(
  UMAP_1   = Embeddings(obj, "umap")[, 1],
  UMAP_2   = Embeddings(obj, "umap")[, 2],
  iLISI    = iLISI, cLISI = cLISI,
  celltype = obj$celltype)

cat(sprintf("iLISI observed (ideal=%d): median=%.2f mean=%.2f\n", n_samp, median(iLISI), mean(iLISI)))
cat(sprintf("iLISI shuffled null (achievable ceiling): median=%.2f\n", null_ceil))
cat(sprintf("observed / ceiling = %.2f  (1.0 = as mixed as theoretically possible)\n",
            median(iLISI) / null_ceil))
cat(sprintf("cLISI (celltype, ideal=1): median=%.2f mean=%.2f\n", median(cLISI), mean(cLISI)))
```

### 6e. iLISI map — *visualizing* sample mixing

Bright = many effective samples in the local neighbourhood (well mixed); dark =
few (sample-dominated).

```{r ilisi_umap, fig.width = 8, fig.height = 6.5}
ggplot(ldat, aes(UMAP_1, UMAP_2, color = iLISI)) +
  geom_point(size = 0.25) +
  scale_color_viridis_c(option = "viridis", name = "iLISI\n(eff. # samples)") +
  theme_classic() +
  labs(title = "Local sample mixing (iLISI, perplexity 30)",
       subtitle = sprintf("median = %.1f of %d samples (higher = better mixed; n=%d nuclei)",
                          median(iLISI), n_samp, length(iLISI)))
```

```{r ilisi_boxplot, fig.width = 8, fig.height = 5}
ggplot(ldat, aes(reorder(celltype, iLISI, median), iLISI, fill = celltype)) +
  geom_boxplot(outlier.size = 0.2) +
  geom_hline(yintercept = null_ceil, linetype = "dashed", color = "red") +
  coord_flip() + theme_bw() + NoLegend() +
  labs(x = NULL, y = "iLISI (effective # samples)",
       title = "Sample mixing per cell type",
       subtitle = sprintf("red dashed = achievable ceiling (shuffled null = %.1f); floor = 1", null_ceil))
```

### 6f. Observed vs label-shuffled null — mixing is near the achievable ceiling

```{r ilisi_null_compare, fig.width = 6, fig.height = 5}
cdat <- rbind(
  data.frame(group = "observed", iLISI = iLISI),
  data.frame(group = "shuffled null", iLISI = iLISI_null))

ggplot(cdat, aes(group, iLISI, fill = group)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  geom_hline(yintercept = 1, linetype = "dotted") +
  annotate("text", x = 0.6, y = 1.6, label = "floor = 1 (no mixing)", size = 3, hjust = 0) +
  scale_fill_manual(values = c(observed = "#0072B2", `shuffled null` = "grey70")) +
  theme_bw() + NoLegend() +
  labs(x = NULL, y = "iLISI (effective # samples)",
       title = "Observed mixing vs achievable ceiling",
       subtitle = sprintf("observed median %.1f = %.0f%% of shuffled ceiling %.1f (ideal 47 is unreachable)",
                          median(iLISI), 100 * median(iLISI) / null_ceil, null_ceil))
```

---

## Session info
```{r session}
sessionInfo()
```
